---
title: "A McKendrick-von Foerster Age-Structured SIRS Model With Immunity in R"
author: ""
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5,
  fig.retina = 2
)

library(deSolve)
library(ReacTran)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

theme_set(theme_minimal(base_size = 13))
```

```{r grids-and-parameters}
age_max <- 80 # Setting oldest age in population to 80 years [MAY WANT TO CHANGE THIS?]
da <- 0.25 # Width of one age bin in years
dt <- 0.1 # Size of one time step in years
time_max <- 12 # Simulation horizon in years

grid <- setup.grid.1D(x.up = 0, L = age_max, N = round(age_max / da))
age <- grid$x.mid # Age values at bin centers
time <- seq(0, time_max, by = dt)

n_age <- length(age) # number of age bins
cfl <- dt / da 
# Set CFL condition: Do not let one numerical time step age people by more than about one full age bin. 
stopifnot(cfl <= 1) # sanity check 

#Population scale, convert relative densities to pop counts
N_target <- 200000  # Total population size 200,000 people, approximate population of Navajo Nation in 1988

# Age distribution/shape
N0_shape <- exp(-0.035 * age) 
N0_shape <- N0_shape / sum(N0_shape * da)  # Normalized baseline density (integrates to 1)

# Population density scaled to N_target persons
N0_density <- N0_shape * N_target

beta_matrix <- matrix(c( # Reference Jackson et al. 2012, Appendix 2 Table 2, contact matrix for Alaska Native population)
  # Row: Age of susceptible population | # Column: Age of infectious population
  # [0,1] [2,4] [5,9] [>=10]
    2.11, 0.15, 0.53, 0.03,  #[0,1]
    0.55, 0.40, 0.50, 0.12,  #[2,4]
    0.56, 3.68, 3.61, 0.13,  #[5,9]
    0.55, 0.55, 0.81, 1.43   #[>=10]
), nrow=4, byrow=TRUE) # beta_ij
age_group_idx <- dplyr::case_when(
  age < 2.0 ~ 1,
  age >= 2.0 & age < 5.0 ~ 2,
  age >= 5.0 & age < 10.0 ~ 3,
  TRUE ~ 4
)

susceptibility <- rep(1, length(age)) # constant for now, account for age and immunity level later / need to check with Shaun/Josh
infectiousness <- rep(1, length(age)) # constant for now, account for age and immunity level later / need to check with Shaun/Josh


mortality <- 0.008 + 0.06 / (1 + exp(-(age - 72) / 5))
# Logistic (sigmoid curve) with background mortality of 0.008, steep increase starting around age 72, and width of about 5 years. This is just an illustrative shape for now. We could also do a more realistic U-shaped curve with high mortality in infancy and old age, but this is simpler for now.

fertility_raw <- exp(-0.5 * ((age - 22) / 6)^2)
# Pure Gaussian shape with peak fertility of 1 at age 22 and width of about 6 years (common fertility curve adjusted to 1988 Navajo data)

fertility_scale <- sum(mortality * N0_shape * da) / sum(fertility_raw * N0_shape * da)
# This scaling factor ensures that the initial population is at demographic equilibrium, meaning that the total number of births balances the total number of deaths. We compute it by taking the ratio of the total mortality (integrated over age) to the total raw fertility (also integrated over age), both weighted by the initial age distribution. This way, when we apply the fertility profile to the initial population, we get a birth influx that exactly offsets the deaths, keeping the population stable at the start of the simulation.

fertility <- fertility_scale * fertility_raw
# The final fertility profile is the raw Gaussian shape scaled by the factor computed above to ensure demographic equilibrium at the start of the simulation.

# NEW: Age-dependent rate of progression from carriage to invasive disease, sigma(a)
## Aligned with 4-month maternal antibody clearance & 4-12 month peak vulnerability and reflecting the epidemiology of invasive Hib disease being concentrated in young children.

# Parameters

# Updated age-dependent sigma from Jackson et al. (for Alaska Native population)
# Jackson et al baseline and amplitude are weekly rates, convert to annual rates (* 52.1775):
sigma_base_annual <- 3.50e-5 * 52.1775   # ~0.001826 / year
sigma_amp_annual  <- 0.01591 * 52.1775   # ~0.83014  / year
decay_rate_weekly <- 0.0371              # Per week decay rate

age_in_weeks <- age * 52.1775

# Maternal antibody fractions; proportion of newborns that enter in each immunity status. Currently illustrative. Also need to fix so that rate at which maternal antibodies wane is faster than normal antibody waning effect. 
#frac_maternal_high <- 0.79
#frac_maternal_low <- 0.00
#frac_maternal_none <- 1 - frac_maternal_high - frac_maternal_low

# Empirical maternal antibody clearance rate (half-life ~ 6-8 weeks)
mu_mat <- 0.065 # per week

# Maternal susceptibility multiplier: 0 at birth, approaching 1 as antibodies decay
maternal_decay <- 1 - exp(-mu_mat * age_in_weeks)

# Jackson et al. sigma scaled by maternal antibody loss
sigma <- (sigma_base_annual + sigma_amp_annual * exp(-decay_rate_weekly * age_in_weeks)) * maternal_decay

# DEFINING IMMUNITY CLASSES AND KERNEL
immunity_classes <- c("none", "low", "high")
n_immunity <- length(immunity_classes)

# Susceptibility Multiplier by immunity class (kernel); how vulnerable someone in each class is to getting infected relative to a fully susceptible person
s_immunity <- c(none = 1.0, low = 1.0, high = 0.04)
alpha_L <- 1 - s_immunity["low"] # efficacy of low antibody against carriage acquisition
alpha_H <- 1 - s_immunity["high"] # efficacy of high antibody against carriage acquisition
#LOW s_immunity set to be 1.0 per Jackson et al. (low antibody does not protect against carriage, only high antibody) and HIGH s_immunity set to 0.04 per Jackson et al.

# NEW: efficacy of antibody against progression from carriage to invasive disease.
# These are conceptually distinct from alpha_L/alpha_H (protection against acquiring
# carriage) - illustrative values only, set higher than the carriage-efficacies since
# antibody is generally more protective against invasive disease than against carriage.
beta_L <- 0.90  # efficacy of low antibody in preventing invasive disease among carriers
beta_H <- 0.98  # efficacy of high antibody in preventing invasive disease among carriers
#updated to 0.90 and 0.98 for low and high antibody, respectively, protection against invasive disease

#NEW WANING RATE CODE, per Jackson et al.
# High to Low antibody waning depends on host age:
# <1 year: 4.9987/yr | 1-2 years: 0.3287/yr | >=2 years: 0.1983/yr
wane_high_to_low <- dplyr::case_when(
  age < 1.0 ~ 4.9987,
  age < 2.0 ~ 0.3287,
  TRUE      ~ 0.1983
)
# Constant waning rate for Low to No antibody waning across all ages
wane_low_to_none <- 0.0939

rho_C <- 2.1759 # rate of recovery from carriage (1/duration of carriage episode)
# Based on duration of carriage reported by Jackson et al.
rho_I <- 33.2067 # rate of recovery from invasive disease (1/duration of illness)
# Based on duration of carriage reported by Jackson et al. 

# External Force of Infection
lambda_ext <- 0.0005

# Vaccine Efficacy Baseline (for subsequent policy/vaccination adjustments/simulations)
vaccine_efficacy <- ifelse(age < 1.0, 0.99, 1.00)
```

```{r ingredient-plots}
ingredient_df <- tibble(
  age = age,
  mortality = mortality,
  fertility = fertility,
  sigma = sigma
) |>
  pivot_longer(-age, names_to = "quantity", values_to = "value")

ggplot(ingredient_df, aes(age, value)) +
  geom_line(linewidth = 1.0, colour = "#1f78b4") +
  facet_wrap(~ quantity, scales = "free_y", ncol = 2) +
  labs(
    title = "Age-Dependent Biological & Demographic Inputs",
    x = "Age (years)", y = "Value"
  )
```

# Initial Condition

```{r new-initial-condition}
# CALIBRATED INITIAL CONDITIONS based on Empirical Pre-Vaccine Hib Data from Navajo, White Mountain Apache, and Alaska Native populations)

# 1. Age-varying Maternal Antibody Decay
# 79% of newborns start with high maternal IgG, clearing rapidly by 4 months (~0.33 yrs)
frac_maternal_high <- 0.79
frac_maternal_low  <- 0.00
frac_maternal_none <- 1 - frac_maternal_high - frac_maternal_low

# Steeper decay: at age = 0.33 (4 months), exp(-0.33/0.12) ~ 0.06 (over 90% lost)
frac_high_by_age <- frac_maternal_high * exp(-age / 0.12)
frac_low_by_age  <- frac_maternal_low  * exp(-age / 0.12) + 0.10 * exp(-((age - 1.5) / 1.0)^2)
frac_none_by_age <- pmax(1 - frac_high_by_age - frac_low_by_age, 0)


# 2. Empirical Baseline Carriage Prevalence Function P_C(a)
# - Michaels et al. (1976): ~0.7% under 6 months (0.5 yrs), 3-5% across childhood
# - Singleton (2006) / Millar (2000): ~1.5% in ages 1-4, up to 3.4-5.0% in older childhood
carriage_prob <- dplyr::case_when(
  age < 0.5 ~ 0.007,                                    # < 6 months (~0.7%)
  age >= 0.5 & age < 5.0 ~ 0.015 + 0.020 * (age / 5),  # 0.5 - 5 yrs (1.5% -> 3.5%)
  age >= 5.0 & age < 10.0 ~ 0.034,                      # 5 - 9 yrs (3.4%)
  TRUE ~ 0.010                                          # Adults (~1.0%)
)

# Total Carriage Density across age
C0_total <- N0_density * carriage_prob
D0       <- 0.002 * N0_density * exp(-0.5 * ((age - 0.75) / 0.35)^2)
S0_total <- pmax(N0_density - C0_total - D0, 0)

S0 <- matrix(0, nrow = n_age, ncol = n_immunity, dimnames = list(NULL, immunity_classes))
C0 <- matrix(0, nrow = n_age, ncol = n_immunity, dimnames = list(NULL, immunity_classes))

S0[, "none"] <- S0_total * frac_none_by_age
S0[, "low"]  <- S0_total * frac_low_by_age
S0[, "high"] <- S0_total * frac_high_by_age

C0[, "none"] <- C0_total * frac_none_by_age
C0[, "low"]  <- C0_total * frac_low_by_age
C0[, "high"] <- C0_total * frac_high_by_age

# Plot Initial Distribution in Absolute Counts per Bin
initial_counts_df <- tibble(
  age = age,
  S_none = S0[, "none"] * da,
  S_low  = S0[, "low"]  * da,
  S_high = S0[, "high"] * da,
  C_none = C0[, "none"] * da,
  C_low  = C0[, "low"]  * da,
  C_high = C0[, "high"] * da,
  D      = D0 * da
) |>
  pivot_longer(-age, names_to = "compartment", values_to = "count")

ggplot(initial_counts_df, aes(age, count, colour = compartment)) +
  geom_line(linewidth = 1.0) +
  scale_colour_manual(values = c(
    S_none = "#1b9e77", S_low = "#66c2a5", S_high = "#a6d854",
    C_none = "#fdae61", C_low = "#f46d43", C_high = "#d73027",
    D      = "#7570b3"
  )) +
  labs(
    title = paste0("Initial Population Counts per ", da, "-Year Age Bin"),
    x = "Age (years)", y = "Number of Individuals"
  )

```

# Core Numerical Pieces

## Force of infection and birth influx
The force of infection is now driven by **total carriers**, and the population total in the denominator now includes `C` and `D` as well as `S`.

```{r forcing-functions-and-reaction-step}
force_of_infection <- function(S, C, D, beta_matrix, susceptibility, infectiousness, age_group_idx, da, lambda_ext = 0.0005) {
  total_S <- rowSums(S)
  total_C <- rowSums(C)
  total_pop_by_age <- total_S + total_C + D
  
  # Calculate infectious density and total population mass per fine bin
  inf_bin <- infectiousness * total_C * da
  pop_bin <- total_pop_by_age * da
  
  # Initialize group vectors for all 4 groups (ensures fixed length 4 even if a group is empty)
  C_group <- numeric(4)
  N_group <- numeric(4)
  
  # Aggregate mass into the 4 contact age groups
  c_agg <- tapply(inf_bin, age_group_idx, sum)
  n_agg <- tapply(pop_bin, age_group_idx, sum)
  
  C_group[as.numeric(names(c_agg))] <- c_agg
  N_group[as.numeric(names(n_agg))] <- n_agg
  
  # Calculate carrier proportion safely (avoid 0/0 -> NaN)
  prop_inf_group <- ifelse(N_group > 0, C_group / N_group, 0)
  
  # Matrix multiplication to get group-level FOI (returns a 4x1 vector)
  lambda_group <- as.vector(beta_matrix %*% prop_inf_group)
  
  # Map back to 160 fine age bins and apply susceptibility modifier
  lambda_age <- lambda_group[age_group_idx] * susceptibility
  
  # Ensure no residual NAs and add external FOI
  lambda_age <- ifelse(is.na(lambda_age), 0, lambda_age) + lambda_ext
  
  return(lambda_age)
}

birth_influx <- function(S, C, D, fertility, da) {
  total_S <- rowSums(S)
  total_C <- rowSums(C)
  sum(fertility * (total_S + total_C + D) * da)
}

age_transport_step <- function(u, boundary_value, dt, grid) {
  transport_change <- tran.1D(
    C = u, C.up = boundary_value, C.down = tail(u, 1),
    v = 1, D = 0, dx = grid
  )$dC
  u + dt * transport_change
}

age_transport_step_matrix <- function(M, boundary_values, dt, grid) {
  out <- M
  for (j in seq_len(ncol(M))) {
    out[, j] <- age_transport_step(M[, j], boundary_values[j], dt, grid)
  }
  out
}

pack_state <- function(S, C, D) {
  c(as.vector(S), as.vector(C), D)
}

unpack_state <- function(y, n_age, n_immunity, immunity_classes) {
  n_block <- n_age * n_immunity
  S <- matrix(y[1:n_block], nrow = n_age, ncol = n_immunity, dimnames = list(NULL, immunity_classes))
  C <- matrix(y[(n_block + 1):(2 * n_block)], nrow = n_age, ncol = n_immunity, dimnames = list(NULL, immunity_classes))
  D <- y[(2 * n_block + 1):(2 * n_block + n_age)]
  list(S = S, C = C, D = D)
}

reaction_rhs <- function(t, y, parms) {
  with(parms, {
    state <- unpack_state(y, n_age, n_immunity, immunity_classes)
    
    S_none <- state$S[, "none"]
    S_low  <- state$S[, "low"]
    S_high <- state$S[, "high"]
    C_none <- state$C[, "none"]
    C_low  <- state$C[, "low"]
    C_high <- state$C[, "high"]
    D      <- state$D
    
    lambda_none <- lambda 
    lambda_low  <- lambda * (1 - alpha_L)
    lambda_high <- lambda * (1 - alpha_H)
    
    sigma_none <- sigma
    sigma_low  <- sigma * (1 - beta_L)
    sigma_high <- sigma * (1 - beta_H)
    
    dS_none <- -(lambda_none + mortality) * S_none + wane_low_to_none * S_low
    dS_low  <- -(lambda_low + wane_low_to_none + mortality) * S_low + wane_high_to_low * S_high 
    dS_high <- -(lambda_high + wane_high_to_low + mortality) * S_high + (C_none + C_low + C_high) * rho_C + rho_I * D
    
    dC_none <- lambda_none * S_none - (sigma_none + rho_C + mortality) * C_none
    dC_low  <- lambda_low * S_low - (sigma_low + rho_C + mortality) * C_low
    dC_high <- lambda_high * S_high - (sigma_high + rho_C + mortality) * C_high
    
    disease_inflow <- sigma_none * C_none + sigma_low * C_low + sigma_high * C_high
    dD             <- disease_inflow - (rho_I + mortality) * D
    
    dS <- cbind(none = dS_none, low = dS_low, high = dS_high)
    dC <- cbind(none = dC_none, low = dC_low, high = dC_high)
    
    list(c(as.vector(dS), as.vector(dC), dD))
  })
}

reaction_step_desolve <- function(
  S_star, C_star, D_star, lambda, sigma, mortality, dt, n_age, n_immunity, 
  immunity_classes, alpha_L, alpha_H, beta_L, beta_H, 
  wane_high_to_low, wane_low_to_none, rho_C, rho_I
) {
  y_star <- pack_state(S_star, C_star, D_star)

  sol <- ode(
    y = y_star, times = c(0, dt), func = reaction_rhs,
    parms = list(
      lambda = lambda, sigma = sigma, mortality = mortality,
      n_age = n_age, n_immunity = n_immunity, immunity_classes = immunity_classes,
      alpha_L = alpha_L, alpha_H = alpha_H, beta_L = beta_L, beta_H = beta_H,
      wane_high_to_low = wane_high_to_low, wane_low_to_none = wane_low_to_none,
      rho_C = rho_C, rho_I = rho_I
    ),
    method = "radau", rtol = 1e-6, atol = 1e-8
  )

  unpack_state(sol[2, -1], n_age, n_immunity, immunity_classes)
}

```


```{r simulator}
simulate_mcvf_sir_imex <- function(
  S0, C0, D0, age, time, da, dt, grid, beta_matrix, age_group_idx, sigma,
  susceptibility, infectiousness, fertility, mortality, 
  n_immunity, immunity_classes, alpha_L, alpha_H, beta_L, beta_H, 
  wane_high_to_low, wane_low_to_none, rho_C, rho_I,
  frac_maternal_none, frac_maternal_low, frac_maternal_high,
  snapshot_times = c(0, 1, 2, 4, 8, 12), heatmap_stride = 2
) {
  S <- S0
  C <- C0
  D <- D0
  
  n_age <- length(age)
  summary_store  <- vector("list", length(time))
  snapshot_store <- list()
  heatmap_store  <- list()
  snap_counter   <- 1

  for (n in seq_along(time)) {
    t_now <- time[n]

    # UPDATED FOI CALL
    lambda_now <- force_of_infection(
      S = S, C = C, D = D, 
      beta_matrix = beta_matrix, 
      susceptibility = susceptibility, 
      infectiousness = infectiousness, 
      age_group_idx = age_group_idx, 
      da = da, 
      lambda_ext = lambda_ext
    )
    
    births_now <- birth_influx(S, C, D, fertility, da)

    total_S_by_class <- colSums(S * da)
    total_S          <- sum(total_S_by_class)
    
    total_C_by_class <- colSums(C * da)
    total_C          <- sum(total_C_by_class)
    
    total_D <- sum(D * da)
    total_N <- total_S + total_C + total_D
    
    carriage_incidence_now <- sum(lambda_now * rowSums(S) * da)
    disease_incidence_now  <- sum(sigma * rowSums(C) * da)

    summary_store[[n]] <- tibble(
      time = t_now, S = total_S,
      S_none = total_S_by_class["none"], S_low = total_S_by_class["low"], S_high = total_S_by_class["high"],
      C = total_C,
      C_none = total_C_by_class["none"], C_low = total_C_by_class["low"], C_high = total_C_by_class["high"],
      D = total_D, N = total_N, births = births_now,
      carriage_incidence = carriage_incidence_now, disease_incidence = disease_incidence_now,
      mean_lambda = mean(lambda_now)
    )

    if (any(abs(t_now - snapshot_times) < dt / 2)) {
      snapshot_store[[snap_counter]] <- tibble(
        time = t_now, age = age,
        S_none_count = S[, "none"] * da, S_low_count = S[, "low"] * da, S_high_count = S[, "high"] * da,
        C_none_count = C[, "none"] * da, C_low_count = C[, "low"] * da, C_high_count = C[, "high"] * da,
        D_count      = D * da,
        total_carriage_count = rowSums(C) * da,
        total_disease_count  = D * da,
        new_disease_cases_per_year = (sigma * rowSums(C)) * da
      )
      snap_counter <- snap_counter + 1
    }

    if ((n - 1) %% heatmap_stride == 0) {
      heatmap_store[[length(heatmap_store) + 1]] <- tibble(
        time = t_now, age = age,
        D_cases = D * da,
        carriage_cases = rowSums(C) * da,
        prevalence_disease = D / pmax(rowSums(S) + rowSums(C) + D, 1e-12),
        prevalence_carriage = rowSums(C) / pmax(rowSums(S) + rowSums(C) + D, 1e-12)
      )
    }

    if (n == length(time)) break
    
    S_boundary <- c(none = births_now * frac_maternal_none,
                    low  = births_now * frac_maternal_low,
                    high = births_now * frac_maternal_high)[immunity_classes]
    C_boundary <- c(none = 0, low = 0, high = 0)[immunity_classes]
    D_boundary <- 0

    S_star <- age_transport_step_matrix(S, S_boundary, dt, grid)
    C_star <- age_transport_step_matrix(C, C_boundary, dt, grid)
    D_star <- age_transport_step(D, D_boundary, dt, grid)

    react <- reaction_step_desolve(
      S_star = S_star, C_star = C_star, D_star = D_star,
      lambda = lambda_now, sigma = sigma, mortality = mortality, dt = dt,
      n_age = n_age, n_immunity = n_immunity, immunity_classes = immunity_classes,
      alpha_L = alpha_L, alpha_H = alpha_H, beta_L = beta_L, beta_H = beta_H,
      wane_high_to_low = wane_high_to_low, wane_low_to_none = wane_low_to_none,
      rho_C = rho_C, rho_I = rho_I
    )

    S <- pmax(react$S, 0)
    C <- pmax(react$C, 0)
    D <- pmax(react$D, 0)
  }

  list(
    summary   = bind_rows(summary_store),
    snapshots = bind_rows(snapshot_store),
    heatmap   = bind_rows(heatmap_store)
  )
}
```

# Run the Simulation

```{r run-simulation}
out <- simulate_mcvf_sir_imex(
  S0 = S0, C0 = C0, D0 = D0, age = age, time = time, da = da, dt = dt, grid = grid,
  beta_matrix = beta_matrix, age_group_idx = age_group_idx, sigma = sigma, 
  susceptibility = susceptibility, infectiousness = infectiousness,
  fertility = fertility, mortality = mortality, n_immunity = n_immunity,
  immunity_classes = immunity_classes, alpha_L = alpha_L, alpha_H = alpha_H,
  beta_L = beta_L, beta_H = beta_H, wane_high_to_low = wane_high_to_low,
  wane_low_to_none = wane_low_to_none, rho_C = rho_C, rho_I = rho_I,
  frac_maternal_none = frac_maternal_none, frac_maternal_low = frac_maternal_low,
  frac_maternal_high = frac_maternal_high, snapshot_times = c(0, 1, 2, 4, 8, 12),
  heatmap_stride = 2
)
summary_df   <- out$summary
snapshots_df <- out$snapshots
heatmap_df   <- out$heatmap

kable(head(summary_df, 8), digits = 2)

```


#pop totals over time
```{r pop-totals}
summary_long <- summary_df |>
  select(time, S, C, D) |>
  pivot_longer(-time, names_to = "compartment", values_to = "persons")

ggplot(summary_long, aes(time, persons, colour = compartment)) +
  geom_line(linewidth = 1.0) +
  scale_colour_manual(values = c(S = "#1b9e77", C = "#d95f02", D = "#7570b3")) +
  labs(
    title = paste0("Total Population Mass Across Compartments (N = ", format(N_target, big.mark=","), ")"),
    x = "Time (years)", y = "Total Persons"
  )

```

#incidence counts
```{r incidence-counts}
summary_aux <- summary_df |>
  select(time, carriage_incidence, disease_incidence, births) |>
  pivot_longer(-time, names_to = "quantity", values_to = "value")

ggplot(summary_aux, aes(time, value)) +
  geom_line(linewidth = 1.0, colour = "#2c7fb8") +
  facet_wrap(~ quantity, scales = "free_y", ncol = 1) +
  labs(
    title = "Annualized Absolute Incidence Counts & Births",
    x = "Time (years)", y = "Persons / Year"
  )

```

#age-specific counts over time
```{r age-counts}
snapshot_counts_long <- snapshots_df |>
  pivot_longer(
    cols = c(S_none_count, S_low_count, S_high_count, 
             C_none_count, C_low_count, C_high_count, D_count),
    names_to = "compartment", values_to = "count"
  ) |>
  mutate(
    compartment = gsub("_count", "", compartment),
    time_label  = paste0("t = ", format(round(time, 1), nsmall = 1))
  )

ggplot(snapshot_counts_long, aes(age, count, colour = compartment)) +
  geom_line(linewidth = 0.9) +
  facet_grid(compartment ~ time_label, scales = "free_y") +
  scale_colour_manual(values = c(
    S_none = "#1b9e77", S_low = "#66c2a5", S_high = "#a6d854",
    C_none = "#fdae61", C_low = "#f46d43", C_high = "#d73027",
    D      = "#7570b3"
  )) +
  labs(
    title = paste0("Age Profiles by Compartment (Counts per ", da, "-Year Bin)"),
    x = "Age (years)", y = "Number of Individuals"
  ) +
  theme(legend.position = "none")

```

#childhood (<5 y-o) invasive disease cases
```{r child-cases}
snapshots_df |>
  filter(age <= 5) |>
  mutate(time_label = paste0("t = ", format(round(time, 1), nsmall = 1))) |>
  ggplot(aes(age, new_disease_cases_per_year, colour = time_label)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "Annualized New Invasive Hib Cases by Age (<5 Years)",
    subtitle = "Peak vulnerability driven by maternal IgG decay at 4 months",
    x = "Age (years)", y = "New Cases / Year (per Bin)", colour = "Time Step"
  )

```

#cases heatmap (absolute)
```{r heatmap-cases}
ggplot(heatmap_df, aes(time, age, fill = D_cases)) +
  geom_raster(interpolate = FALSE) +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b") +
  labs(
    title = paste0("Invasive Disease Cases (D) per ", da, "-Year Age Bin"),
    x = "Time (years)", y = "Age (years)", fill = "Cases"
  )

ggplot(heatmap_df, aes(time, age, fill = prevalence_disease)) +
  geom_raster(interpolate = FALSE) +
  scale_fill_gradient(low = "#fff5eb", high = "#7f2704") +
  labs(
    title = "Age-Specific Invasive Disease Prevalence Profile",
    x = "Time (years)", y = "Age (years)", fill = "Prevalence"
  )

```


# Aggregate Dynamics

These are the total susceptible, infectious, and recovered masses after integrating over age.

```{r aggregate-dynamics}
summary_long <- summary_df |>
  select(time, S, C, D) |>
  pivot_longer(-time, names_to = "compartment", values_to = "mass")

ggplot(summary_long, aes(time, mass, colour = compartment)) +
  geom_line(linewidth = 1.0) +
  scale_colour_manual(values = c(S = "#1b9e77", C = "#d95f02", D = "#7570b3")) +
  labs(
    title = "Aggregate SCD dynamics",
    x = "Time (years)",
    y = "Total mass"
  )
```

```{r incidence-and-births}
summary_aux <- summary_df |>
  select(time, carriage_incidence, disease_incidence, births, mean_lambda) |>
  pivot_longer(-time, names_to = "quantity", values_to = "value")

ggplot(summary_aux, aes(time, value)) +
  geom_line(linewidth = 1.0, colour = "#2c7fb8") +
  facet_wrap(~ quantity, scales = "free_y", ncol = 1) +
  labs(
    title = "Carriage/disease incidence, births, and mean force of infection",
    x = "Time (years)",
    y = NULL
  )
```

# Age Profiles at Selected Times

```{r age-profiles}
snapshot_long <- snapshots_df |>
  pivot_longer(
    cols = c(S_none_count, S_low_count, S_high_count,
             C_none_count, C_low_count, C_high_count, D_count),
    names_to = "compartment", values_to = "density"
  ) |>
  mutate(
    compartment = gsub("_count", "", compartment),
    time_label  = paste0("t = ", format(round(time, 1), nsmall = 1))
  )

ggplot(snapshot_long, aes(age, density, colour = compartment)) +
  geom_line(linewidth = 0.9) +
  facet_grid(compartment ~ time_label, scales = "free_y") +
  scale_colour_manual(values = c(
  S_none = "#1b9e77", S_low = "#66c2a5", S_high = "#a6d854",
    C_none = "#fdae61", C_low = "#f46d43", C_high = "#d73027",
    D = "#7570b3"
)) +
  labs(
    title = "Age profiles of S, C, and D at selected times",
    x = "Age (years)",
    y = "Density"
  ) +
  theme(legend.position = "none")
```

# Heatmaps

These heatmaps are often the easiest way to see what an age-structured McVF model is doing.

```{r disease-heatmap}
ggplot(heatmap_df, aes(time, age, fill = D_cases)) +
  geom_raster(interpolate = FALSE) +
  scale_fill_gradient(low = "#f7fbff", high = "#08306b") +
  labs(
    title = "Invasive disease density over age and time",
    x = "Time (years)",
    y = "Age (years)",
    fill = "D"
  )
```

```{r carriage-heatmap}
ggplot(heatmap_df, aes(time, age, fill = carriage_cases)) +
  geom_raster(interpolate = FALSE) +
  scale_fill_gradient(low = "#fff5eb", high = "#7f2704") +
  labs(
    title = "Carriage density over age and time",
    x = "Time (years)",
    y = "Age (years)",
    fill = "C"
  )
```


```{r prevalence-heatmap}
prevalence_long <- heatmap_df |>
  select(time, age, prevalence_disease, prevalence_carriage) |>
  pivot_longer(c(prevalence_disease, prevalence_carriage), names_to = "quantity", values_to = "value")

ggplot(prevalence_long, aes(time, age, fill = value)) +
  geom_raster(interpolate = FALSE) +
  scale_fill_gradient(low = "#fff5eb", high = "#7f2704") +
  facet_wrap(~ quantity) +
  labs(
    title = "Age-specific prevalence over age and time",
    x = "Time (years)",
    y = "Age (years)",
    fill = "Prev."
  )
```

