---
title: "A McKendrick-von Foerster Age-Structured SIRS Model With Immunity in R"
author: ""
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 9,
  fig.height = 5.5,
  fig.retina = 2
)

library(deSolve)
library(ReacTran)
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)

theme_set(theme_minimal(base_size = 13))
```

```{r grids-and-parameters}
age_max <- 80 # Setting oldest age in population to 80 years
da <- 0.25    # Width of one age bin in years
dt <- 0.1     # Size of one time step in years
time_max <- 20 # Baseline simulation horizon in years

grid <- setup.grid.1D(x.up = 0, L = age_max, N = round(age_max / da))
age <- grid$x.mid # Age values at bin centers
time <- seq(0, time_max, by = dt)

n_age <- length(age) # Number of age bins
cfl <- dt / da 
stopifnot(cfl <= 1) # Sanity check for CFL condition

# Population scale
N_target <- 200000 

# Age distribution/shape
N0_shape <- exp(-0.035 * age) 
N0_shape <- N0_shape / sum(N0_shape * da)

# Population density scaled to N_target persons
N0_density <- N0_shape * N_target

beta_matrix <- matrix(c(
    2.11, 0.15, 0.53, 0.03,  #[0,1]
    0.55, 0.40, 0.50, 0.12,  #[2,4]
    0.56, 3.68, 3.61, 0.13,  #[5,9]
    0.55, 0.55, 0.81, 1.43   #[>=10]
), nrow = 4, byrow = TRUE)

age_group_idx <- dplyr::case_when(
  age < 2.0 ~ 1,
  age >= 2.0 & age < 5.0 ~ 2,
  age >= 5.0 & age < 10.0 ~ 3,
  TRUE ~ 4
)

susceptibility <- rep(1, length(age))
infectiousness <- rep(1, length(age))

# ==============================================================================
# DEMOGRAPHIC TIME-VARYING MATRICES
# ==============================================================================
demo_time <- seq(0, 50, by = 1)
n_demo_time <- length(demo_time)

mortality_base <- 0.008 + 0.06 / (1 + exp(-(age - 72) / 5))
fertility_raw_base <- exp(-0.5 * ((age - 22) / 6)^2)

mortality_decline <- 0.99^(demo_time)
fertility_drop <- seq(1, 0.7, length.out = n_demo_time)

mortality_mat <- matrix(NA, nrow = n_age, ncol = n_demo_time)
fertility_mat <- matrix(NA, nrow = n_age, ncol = n_demo_time)
fertility_mat_scaled <- matrix(NA, nrow = n_age, ncol = n_demo_time)

for (t in 1:n_demo_time) {
  mortality_mat[, t] <- mortality_base * mortality_decline[t]
  fertility_mat[, t] <- fertility_raw_base * fertility_drop[t]
  
  scale_t <- sum(mortality_mat[, t] * N0_shape * da) / sum(fertility_mat[, t] * N0_shape * da)
  fertility_mat_scaled[, t] <- scale_t * fertility_mat[, t]
}

total_deaths_base <- sum(mortality_mat[, 1] * N0_density * da)
raw_births_base <- sum(fertility_mat_scaled[, 1] * N0_density * da)

fertility_mat_scaled[, 1] <- fertility_mat_scaled[, 1] * (total_deaths_base / raw_births_base)

fertility_mat_burnin <- matrix(fertility_mat_scaled[, 1], nrow = n_age, ncol = n_demo_time)
mortality_mat_burnin <- matrix(mortality_mat[, 1], nrow = n_age, ncol = n_demo_time)

# ==============================================================================
# DISEASE & IMMUNITY PARAMETERS
# ==============================================================================
sigma_base_annual <- 3.50e-5 * 52.1775
sigma_amp_annual  <- 0.01591 * 52.1775
decay_rate_weekly <- 0.0371

age_in_weeks <- age * 52.1775
mu_mat <- 0.065
maternal_decay <- 1 - exp(-mu_mat * age_in_weeks)

sigma <- (sigma_base_annual + sigma_amp_annual * exp(-decay_rate_weekly * age_in_weeks)) * maternal_decay

immunity_classes <- c("none", "low", "high")
n_immunity <- length(immunity_classes)

s_immunity <- c(none = 1.0, low = 1.0, high = 0.04)
alpha_L <- 1 - s_immunity["low"]
alpha_H <- 1 - s_immunity["high"]

beta_L <- 0.90
beta_H <- 0.98

wane_high_to_low <- dplyr::case_when(
  age < 1.0 ~ 4.9987,
  age < 2.0 ~ 0.3287,
  TRUE      ~ 0.1983
)
wane_low_to_none <- 0.0939

rho_C <- 2.1759
rho_I <- 33.2067

lambda_ext <- 0.0005
vaccine_efficacy <- ifelse(age < 1.0, 0.99, 1.00)

# Define vaccination timeline parameters
time_vax <- seq(0, 30, by = dt)
n_time_steps <- length(time_vax)
```

```{r ingredient-plots}
ingredient_df <- tibble(
  age = age,
  mortality = mortality_mat[, 1],
  fertility = fertility_mat_scaled[, 1],
  sigma = sigma
) |>
  pivot_longer(-age, names_to = "quantity", values_to = "value")

ggplot(ingredient_df, aes(age, value)) +
  geom_line(linewidth = 1.0, colour = "#1f78b4") +
  facet_wrap(~ quantity, scales = "free_y", ncol = 2) +
  labs(
    title = "Age-Dependent Biological & Demographic Inputs",
    x = "Age (years)", y = "Value"
  )

# ==============================================================================
# REALISTIC HIB DOSE SCHEDULE & TIME-VARYING VACCINATION MATRIX
# ==============================================================================

idx_dose1   <- which(age >= 1.5 / 12 & age <= 3.0 / 12)
idx_dose2   <- which(age >= 3.5 / 12 & age <= 5.0 / 12)
idx_booster <- which(age >= 11.5 / 12 & age <= 15.0 / 12)

coverage_trend <- pmin(0.50 + 0.40 * (time_vax / 3)^0.8, 0.90)

v_rate_none_to_low_matrix <- matrix(0, nrow = n_age, ncol = n_time_steps)
v_rate_to_high_matrix      <- matrix(0, nrow = n_age, ncol = n_time_steps)

for (k in seq_len(n_time_steps)) {
  cov_k <- coverage_trend[k]
  
  v_rate_none_to_low_matrix[idx_dose1, k] <- -log(1 - cov_k * 0.65) / (da)
  v_rate_to_high_matrix[idx_dose2, k]     <- -log(1 - cov_k * 0.92) / (da)
  v_rate_to_high_matrix[idx_booster, k]   <- -log(1 - cov_k * 0.95) / (da * 3)
}
```

# Initial Condition

```{r new-initial-condition}
frac_maternal_high <- 0.79
frac_maternal_low  <- 0.00
frac_maternal_none <- 1 - frac_maternal_high - frac_maternal_low

frac_high_by_age <- frac_maternal_high * exp(-age / 0.12)
frac_low_by_age  <- frac_maternal_low  * exp(-age / 0.12) + 0.10 * exp(-((age - 1.5) / 1.0)^2)
frac_none_by_age <- pmax(1 - frac_high_by_age - frac_low_by_age, 0)

carriage_prob <- dplyr::case_when(
  age < 0.5 ~ 0.007,
  age >= 0.5 & age < 5.0 ~ 0.015 + 0.020 * (age / 5),
  age >= 5.0 & age < 10.0 ~ 0.034,
  TRUE ~ 0.010
)

C0_total <- N0_density * carriage_prob
D0       <- 0.002 * N0_density * exp(-0.5 * ((age - 0.75) / 0.35)^2)
S0_total <- pmax(N0_density - C0_total - D0, 0)

S0 <- matrix(0, nrow = n_age, ncol = n_immunity, dimnames = list(NULL, immunity_classes))
C0 <- matrix(0, nrow = n_age, ncol = n_immunity, dimnames = list(NULL, immunity_classes))

S0[, "none"] <- S0_total * frac_none_by_age
S0[, "low"]  <- S0_total * frac_low_by_age
S0[, "high"] <- S0_total * frac_high_by_age

C0[, "none"] <- C0_total * frac_none_by_age
C0[, "low"]  <- C0_total * frac_low_by_age
C0[, "high"] <- C0_total * frac_high_by_age

initial_counts_df <- tibble(
  age = age,
  S_none = S0[, "none"] * da,
  S_low  = S0[, "low"]  * da,
  S_high = S0[, "high"] * da,
  C_none = C0[, "none"] * da,
  C_low  = C0[, "low"]  * da,
  C_high = C0[, "high"] * da,
  D      = D0 * da
) |>
  pivot_longer(-age, names_to = "compartment", values_to = "count")

ggplot(initial_counts_df, aes(age, count, colour = compartment)) +
  geom_line(linewidth = 1.0) +
  scale_colour_manual(values = c(
    S_none = "#1b9e77", S_low = "#66c2a5", S_high = "#a6d854",
    C_none = "#fdae61", C_low = "#f46d43", C_high = "#d73027",
    D      = "#7570b3"
  )) +
  labs(
    title = paste0("Initial Population Counts per ", da, "-Year Age Bin"),
    x = "Age (years)", y = "Number of Individuals"
  )
```

# Core Numerical Pieces

## Force of infection and birth influx
The force of infection is now driven by **total carriers**, and the population total in the denominator now includes `C` and `D` as well as `S`.

```{r forcing-functions-and-reaction-step}
force_of_infection <- function(S, C, D, beta_matrix, susceptibility, infectiousness, age_group_idx, da, lambda_ext = 0.0005) {
  total_S <- rowSums(S)
  total_C <- rowSums(C)
  total_pop_by_age <- total_S + total_C + D
  
  inf_bin <- infectiousness * total_C * da
  pop_bin <- total_pop_by_age * da
  
  C_group <- numeric(4)
  N_group <- numeric(4)
  
  c_agg <- tapply(inf_bin, age_group_idx, sum)
  n_agg <- tapply(pop_bin, age_group_idx, sum)
  
  C_group[as.numeric(names(c_agg))] <- c_agg
  N_group[as.numeric(names(n_agg))] <- n_agg
  
  prop_inf_group <- ifelse(N_group > 0, C_group / N_group, 0)
  lambda_group <- as.vector(beta_matrix %*% prop_inf_group)
  lambda_age <- lambda_group[age_group_idx] * susceptibility
  lambda_age <- ifelse(is.na(lambda_age), 0, lambda_age) + lambda_ext
  
  return(lambda_age)
}

birth_influx <- function(S, C, D, fertility, da) {
  total_S <- rowSums(S)
  total_C <- rowSums(C)
  sum(fertility * (total_S + total_C + D) * da)
}

age_transport_step <- function(u, boundary_value, dt, grid) {
  transport_change <- tran.1D(
    C = u, C.up = boundary_value, C.down = tail(u, 1),
    v = 1, D = 0, dx = grid
  )$dC
  u + dt * transport_change
}

age_transport_step_matrix <- function(M, boundary_values, dt, grid) {
  out <- M
  for (j in seq_len(ncol(M))) {
    out[, j] <- age_transport_step(M[, j], boundary_values[j], dt, grid)
  }
  out
}

pack_state <- function(S, C, D) {
  c(as.vector(S), as.vector(C), D)
}

unpack_state <- function(y, n_age, n_immunity, immunity_classes) {
  n_block <- n_age * n_immunity
  S <- matrix(y[1:n_block], nrow = n_age, ncol = n_immunity, dimnames = list(NULL, immunity_classes))
  C <- matrix(y[(n_block + 1):(2 * n_block)], nrow = n_age, ncol = n_immunity, dimnames = list(NULL, immunity_classes))
  D <- y[(2 * n_block + 1):(2 * n_block + n_age)]
  list(S = S, C = C, D = D)
}

reaction_rhs <- function(t, y, parms) {
  with(parms, {
    state <- unpack_state(y, n_age, n_immunity, immunity_classes)
    
    S_none <- state$S[, "none"]
    S_low  <- state$S[, "low"]
    S_high <- state$S[, "high"]
    
    C_none <- state$C[, "none"]
    C_low  <- state$C[, "low"]
    C_high <- state$C[, "high"]
    
    D      <- state$D
    
    lambda_none <- lambda 
    lambda_low  <- lambda * (1 - alpha_L)
    lambda_high <- lambda * (1 - alpha_H)
    
    sigma_none <- sigma
    sigma_low  <- sigma * (1 - beta_L)
    sigma_high <- sigma * (1 - beta_H)
    
    v_none_to_low <- if (!is.null(v_rate_none_to_low)) v_rate_none_to_low else numeric(n_age)
    v_to_high     <- if (!is.null(v_rate_to_high)) v_rate_to_high else numeric(n_age)
    
    dS_none <- -(lambda_none + mortality + v_none_to_low + v_to_high) * S_none + wane_low_to_none * S_low
    dS_low  <- -(lambda_low + wane_low_to_none + mortality + v_to_high) * S_low + wane_high_to_low * S_high + v_none_to_low * S_none
    dS_high <- -(lambda_high + wane_high_to_low + mortality) * S_high + (C_none + C_low + C_high) * rho_C + rho_I * D + v_to_high * (S_none + S_low)
    
    dC_none <- lambda_none * S_none - (sigma_none + rho_C + mortality) * C_none
    dC_low  <- lambda_low * S_low - (sigma_low + rho_C + mortality) * C_low
    dC_high <- lambda_high * S_high - (sigma_high + rho_C + mortality) * C_high
    
    disease_inflow <- sigma_none * C_none + sigma_low * C_low + sigma_high * C_high
    dD             <- disease_inflow - (rho_I + mortality) * D
    
    dS <- cbind(none = dS_none, low = dS_low, high = dS_high)
    dC <- cbind(none = dC_none, low = dC_low, high = dC_high)
    
    dy <- c(as.vector(dS), as.vector(dC), as.vector(dD))
    list(dy)
  })
}

reaction_step_desolve <- function(
  S_star, C_star, D_star, lambda, sigma, mortality, dt, n_age, n_immunity, 
  immunity_classes, alpha_L, alpha_H, beta_L, beta_H, 
  wane_high_to_low, wane_low_to_none, rho_C, rho_I, 
  v_rate_none_to_low = NULL, v_rate_to_high = NULL
) {
  y_star <- pack_state(S_star, C_star, D_star)

  sol <- ode(
    y = y_star, times = c(0, dt), func = reaction_rhs,
    parms = list(
      lambda = lambda, sigma = sigma, mortality = mortality,
      n_age = n_age, n_immunity = n_immunity, immunity_classes = immunity_classes,
      alpha_L = alpha_L, alpha_H = alpha_H, beta_L = beta_L, beta_H = beta_H,
      wane_high_to_low = wane_high_to_low, wane_low_to_none = wane_low_to_none,
      rho_C = rho_C, rho_I = rho_I, 
      v_rate_none_to_low = v_rate_none_to_low, v_rate_to_high = v_rate_to_high
    ),
    method = "radau", rtol = 1e-6, atol = 1e-8
  )

  unpack_state(sol[2, -1], n_age, n_immunity, immunity_classes)
}

simulate_mcvf_sir_imex_dynamic <- function(
  S0, C0, D0, age, time, da, dt, grid, beta_matrix, age_group_idx, sigma,
  susceptibility, infectiousness, fertility_matrix, mortality_matrix, 
  n_immunity, immunity_classes, alpha_L, alpha_H, beta_L, beta_H, 
  wane_high_to_low, wane_low_to_none, rho_C, rho_I,
  frac_maternal_none, frac_maternal_low, frac_maternal_high,
  v_rate_none_to_low_matrix = NULL, v_rate_to_high_matrix = NULL,
  lambda_ext = 0.0005,
  snapshot_times = c(0, 5, 10, 15, 20, 25, 30), heatmap_stride = 2
) {
  S <- S0
  C <- C0
  D <- D0
  
  n_age <- length(age)
  summary_store  <- vector("list", length(time))
  snapshot_store <- list()
  snap_counter   <- 1

  for (n in seq_along(time)) {
    t_now <- time[n]

    year_idx <- pmin(floor(t_now) + 1, if (is.matrix(mortality_matrix)) ncol(mortality_matrix) else 1)

    mortality_now <- if (is.matrix(mortality_matrix)) mortality_matrix[, year_idx] else mortality_matrix
    fertility_now <- if (is.matrix(fertility_matrix)) fertility_matrix[, year_idx] else fertility_matrix
    
    # Corrected indexing by simulation step n
    v_none_to_low_now <- if (!is.null(v_rate_none_to_low_matrix)) v_rate_none_to_low_matrix[, n] else NULL
    v_to_high_now     <- if (!is.null(v_rate_to_high_matrix)) v_rate_to_high_matrix[, n] else NULL

    lambda_now <- force_of_infection(
      S = S, C = C, D = D, 
      beta_matrix = beta_matrix, 
      susceptibility = susceptibility, 
      infectiousness = infectiousness, 
      age_group_idx = age_group_idx, 
      da = da, 
      lambda_ext = lambda_ext
    )
    
    total_deaths_now <- sum(mortality_now * (rowSums(S) + rowSums(C) + D) * da)
    births_now <- total_deaths_now

    total_S_by_class <- colSums(S * da)
    total_S          <- sum(total_S_by_class)
    
    total_C_by_class <- colSums(C * da)
    total_C          <- sum(total_C_by_class)
    
    total_D <- sum(D * da)
    total_N <- total_S + total_C + total_D
    
    carriage_incidence_now <- sum(lambda_now * rowSums(S) * da)
    disease_incidence_now  <- sum(sigma * rowSums(C) * da)

    summary_store[[n]] <- tibble(
      time = t_now, 
      year = 1990 + t_now,
      S = total_S,
      S_none = total_S_by_class["none"], S_low = total_S_by_class["low"], S_high = total_S_by_class["high"],
      C = total_C,
      C_none = total_C_by_class["none"], C_low = total_C_by_class["low"], C_high = total_C_by_class["high"],
      D = total_D, N = total_N, births = births_now,
      carriage_incidence = carriage_incidence_now, disease_incidence = disease_incidence_now,
      mean_lambda = mean(lambda_now)
    )

    if (any(abs(t_now - snapshot_times) < dt / 2)) {
      snapshot_store[[snap_counter]] <- tibble(
        time = t_now, year = 1990 + t_now, age = age,
        S_none_count = S[, "none"] * da, S_low_count = S[, "low"] * da, S_high_count = S[, "high"] * da,
        C_none_count = C[, "none"] * da, C_low_count = C[, "low"] * da, C_high_count = C[, "high"] * da,
        D_count      = D * da,
        total_carriage_count = rowSums(C) * da,
        total_disease_count  = D * da,
        new_disease_cases_per_year = (sigma * rowSums(C)) * da
      )
      snap_counter <- snap_counter + 1
    }

    if (n == length(time)) break
    
    S_boundary <- c(none = births_now * frac_maternal_none,
                    low  = births_now * frac_maternal_low,
                    high = births_now * frac_maternal_high)[immunity_classes]
    C_boundary <- c(none = 0, low = 0, high = 0)[immunity_classes]
    D_boundary <- 0

    S_star <- age_transport_step_matrix(S, S_boundary, dt, grid)
    C_star <- age_transport_step_matrix(C, C_boundary, dt, grid)
    D_star <- age_transport_step(D, D_boundary, dt, grid)

    react <- reaction_step_desolve(
      S_star = S_star, C_star = C_star, D_star = D_star,
      lambda = lambda_now, sigma = sigma, mortality = mortality_now, dt = dt,
      n_age = n_age, n_immunity = n_immunity, immunity_classes = immunity_classes,
      alpha_L = alpha_L, alpha_H = alpha_H, beta_L = beta_L, beta_H = beta_H,
      wane_high_to_low = wane_high_to_low, wane_low_to_none = wane_low_to_none,
      rho_C = rho_C, rho_I = rho_I,
      v_rate_none_to_low = v_none_to_low_now,
      v_rate_to_high = v_to_high_now
    )

    S <- pmax(react$S, 0)
    C <- pmax(react$C, 0)
    D <- pmax(react$D, 0)
  }

  list(
    summary     = bind_rows(summary_store),
    snapshots   = bind_rows(snapshot_store),
    final_state = list(S = S, C = C, D = D)
  )
}
```

# Run the Simulation

```{r run-simulation}
# ==============================================================================
# 1. BURN-IN TO ENDEMIC EQUILIBRIUM (NO VAX)
# ==============================================================================
message("Running 100-year burn-in to establish equilibrium initial conditions...")
burnin_horizon <- 100
time_burnin    <- seq(0, burnin_horizon, by = dt)

out_burnin <- simulate_mcvf_sir_imex_dynamic(
  S0 = S0, C0 = C0, D0 = D0, age = age, time = time_burnin, da = da, dt = dt, grid = grid,
  beta_matrix = beta_matrix, age_group_idx = age_group_idx, sigma = sigma, 
  susceptibility = susceptibility, infectiousness = infectiousness,
  fertility_matrix = fertility_mat_burnin, 
  mortality_matrix = mortality_mat_burnin, 
  n_immunity = n_immunity, immunity_classes = immunity_classes,
  alpha_L = alpha_L, alpha_H = alpha_H, beta_L = beta_L, beta_H = beta_H,
  wane_high_to_low = wane_high_to_low, wane_low_to_none = wane_low_to_none,
  rho_C = rho_C, rho_I = rho_I,
  frac_maternal_none = frac_maternal_none, frac_maternal_low = frac_maternal_low,
  frac_maternal_high = frac_maternal_high, 
  snapshot_times = c(0, 50, 100)
)

S0_eq <- out_burnin$final_state$S
C0_eq <- out_burnin$final_state$C
D0_eq <- out_burnin$final_state$D

# ==============================================================================
# 2. RUN COUNTERFACTUAL SCENARIO (30 YEARS, NO VAX, DYNAMIC DEMOGRAPHICS)
# ==============================================================================
message("Running 30-year counterfactual scenario (1990-2020)...")
out_counterfactual <- simulate_mcvf_sir_imex_dynamic(
  S0 = S0_eq, C0 = C0_eq, D0 = D0_eq, age = age, time = time_vax, da = da, dt = dt, grid = grid,
  beta_matrix = beta_matrix, age_group_idx = age_group_idx, sigma = sigma,
  susceptibility = susceptibility, infectiousness = infectiousness,
  fertility_matrix = fertility_mat_scaled,
  mortality_matrix = mortality_mat,
  n_immunity = n_immunity, immunity_classes = immunity_classes,
  alpha_L = alpha_L, alpha_H = alpha_H, beta_L = beta_L, beta_H = beta_H,
  wane_high_to_low = wane_high_to_low, wane_low_to_none = wane_low_to_none,
  rho_C = rho_C, rho_I = rho_I,
  frac_maternal_none = frac_maternal_none, frac_maternal_low = frac_maternal_low,
  frac_maternal_high = frac_maternal_high,
  v_rate_none_to_low_matrix = NULL,
  v_rate_to_high_matrix = NULL
)

# ==============================================================================
# 3. RUN VACCINATION SCENARIO (30 YEARS, REALISTIC HIB VAX, DYNAMIC DEMOGRAPHICS)
# ==============================================================================
message("Running 30-year Hib vaccination scenario (1990-2020)...")
out_vax <- simulate_mcvf_sir_imex_dynamic(
  S0 = S0_eq, C0 = C0_eq, D0 = D0_eq, age = age, time = time_vax, da = da, dt = dt, grid = grid,
  beta_matrix = beta_matrix, age_group_idx = age_group_idx, sigma = sigma,
  susceptibility = susceptibility, infectiousness = infectiousness,
  fertility_matrix = fertility_mat_scaled,
  mortality_matrix = mortality_mat,
  n_immunity = n_immunity, immunity_classes = immunity_classes,
  alpha_L = alpha_L, alpha_H = alpha_H, beta_L = beta_L, beta_H = beta_H,
  wane_high_to_low = wane_high_to_low, wane_low_to_none = wane_low_to_none,
  rho_C = rho_C, rho_I = rho_I,
  frac_maternal_none = frac_maternal_none, frac_maternal_low = frac_maternal_low,
  frac_maternal_high = frac_maternal_high,
  v_rate_none_to_low_matrix = v_rate_none_to_low_matrix,
  v_rate_to_high_matrix = v_rate_to_high_matrix
)
```


```{r diff-analysis}
# ==============================================================================
# DIFFERENCING ANALYSIS & METRICS COMPUTATION (1990 - 2020)
# ==============================================================================
cfr_hib <- 0.05

summary_counter <- out_counterfactual$summary %>% mutate(scenario = "Counterfactual (No Vax)")
summary_vax     <- out_vax$summary %>% mutate(scenario = "Hib Vaccination Program")

comp_df <- summary_counter %>%
  filter(time < 30) %>%
  select(time, year, N, disease_incidence_counter = disease_incidence) %>%
  inner_join(
    summary_vax %>% filter(time < 30) %>% select(time, disease_incidence_vax = disease_incidence),
    by = "time"
  ) %>%
  mutate(
    annual_cases_counter = disease_incidence_counter * dt,
    annual_cases_vax     = disease_incidence_vax * dt,
    cases_averted_dt     = annual_cases_counter - annual_cases_vax,
    
    annual_deaths_counter = annual_cases_counter * cfr_hib,
    annual_deaths_vax     = annual_cases_vax * cfr_hib,
    deaths_averted_dt     = cases_averted_dt * cfr_hib,
    
    rate_100k_counter = (disease_incidence_counter / N) * 100000,
    rate_100k_vax     = (disease_incidence_vax / N) * 100000
  )

annual_impact <- comp_df %>%
  group_by(year = floor(year)) %>%
  summarize(
    pop_size              = mean(N),
    cases_counter         = sum(annual_cases_counter),
    cases_vax             = sum(annual_cases_vax),
    cases_averted         = sum(cases_averted_dt),
    deaths_counter        = sum(annual_deaths_counter),
    deaths_vax            = sum(annual_deaths_vax),
    deaths_averted        = sum(deaths_averted_dt),
    pct_case_reduction    = (1 - cases_vax / cases_counter) * 100,
    .groups = "drop"
  ) %>%
  mutate(
    cum_cases_averted  = cumsum(cases_averted),
    cum_deaths_averted = cumsum(deaths_averted)
  )

tot_cases_counter <- sum(annual_impact$cases_counter)
tot_cases_vax     <- sum(annual_impact$cases_vax)
tot_cases_averted <- sum(annual_impact$cases_averted)

tot_deaths_counter <- sum(annual_impact$deaths_counter)
tot_deaths_vax     <- sum(annual_impact$deaths_vax)
tot_deaths_averted <- sum(annual_impact$deaths_averted)
overall_pct_red    <- (1 - tot_cases_vax / tot_cases_counter) * 100

cumulative_summary <- tibble(
  Metric = c("Total Severe Cases", "Total Deaths", "Overall Reduction (%)"),
  `Counterfactual Scenario` = c(round(tot_cases_counter), round(tot_deaths_counter), "—"),
  `Hib Vaccination Program` = c(round(tot_cases_vax), round(tot_deaths_vax), "—"),
  `Net Averted / Impact`    = c(round(tot_cases_averted), round(tot_deaths_averted), paste0(round(overall_pct_red, 1), "%"))
)

kable(cumulative_summary, caption = "30-Year Cumulative Impact Analysis (1990–2020)")

```


```{r counterfactual-vs-vaccination-trajectories}
# Plot 1: Annual Severe Disease Cases Trajectory
ggplot() +
  geom_line(data = comp_df, aes(x = year, y = disease_incidence_counter, color = "Counterfactual (No Vax)"), linewidth = 1.2) +
  geom_line(data = comp_df, aes(x = year, y = disease_incidence_vax, color = "Hib Vaccination Program"), linewidth = 1.2) +
  scale_color_manual(values = c("Counterfactual (No Vax)" = "#D95F02", "Hib Vaccination Program" = "#1B9E77")) +
  labs(
    title = "Annual Severe Invasive Hib Cases (1990–2020)",
    subtitle = "Navajo Nation 30-Year Simulation with Dynamic Demographics",
    x = "Calendar Year",
    y = "Annual Severe Disease Cases",
    color = "Scenario"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

plot_cum <- annual_impact %>%
  select(year, cum_cases_averted, cum_deaths_averted) %>%
  pivot_longer(-year, names_to = "metric", values_to = "count") %>%
  mutate(metric = ifelse(metric == "cum_cases_averted", "Cumulative Severe Cases Averted", "Cumulative Deaths Averted"))

ggplot(plot_cum, aes(x = year, y = count, fill = metric)) +
  geom_area(alpha = 0.7, color = "black", linewidth = 0.3, position = "identity") +
  facet_wrap(~ metric, scales = "free_y", ncol = 1) +
  scale_fill_manual(values = c("Cumulative Severe Cases Averted" = "#2B8CBE", "Cumulative Deaths Averted" = "#E6550D")) +
  labs(
    title = "Cumulative Public Health Impact of Hib Vaccination (1990–2020)",
    x = "Calendar Year",
    y = "Cumulative Quantity Averted"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))
```




```{r postvax-plots-one}
# ==============================================================================
# 4. PLOT INCIDENCE IMPACT POST-VACCINATION
# ==============================================================================
summary_vax_df <- out_vax$summary %>%
  mutate(
    total_pop = S_none + S_low + S_high + C_none + C_low + C_high + D,
    disease_rate_per_100k  = (disease_incidence / total_pop) * 100000,
    carriage_rate_per_100k = (carriage_incidence / total_pop) * 100000
  )

ggplot(summary_vax_df, aes(x = time)) +
  geom_line(aes(y = disease_rate_per_100k, color = "Invasive Disease"), linewidth = 1.2) +
  geom_line(aes(y = carriage_rate_per_100k / 100, color = "Carriage (x1/100)"), linewidth = 1, linetype = "dashed") +
  labs(
    title = "Impact of Routine Pediatric Vaccination",
    subtitle = "Simulation initialized from burn-in endemic equilibrium",
    x = "Years Since Vaccination Implementation",
    y = "Incidence Rate per 100,000",
    color = "Metric"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom", plot.title = element_text(face = "bold"))

```

```{r postvax-plots-two}
ggplot(summary_vax_df, aes(x = time, y = disease_rate_per_100k)) +
  geom_line(color = "#D55E00", linewidth = 1.2) +
  labs(
    title = "Impact of Routine Pediatric Vaccination on Severe Invasive Disease",
    subtitle = "Overall population incidence following burn-in equilibrium",
    x = "Years Since Vaccination Implementation",
    y = "Severe Disease Incidence Rate per 100,000"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))

```

```{r postvax-plots-three}
# Calculate incidence per 100,000 specifically for age < 5
under5_vax_df <- out_vax$snapshots %>%
  filter(age < 5) %>%
  group_by(time) %>%
  summarize(
    new_cases_under5 = sum(new_disease_cases_per_year),
    pop_under5       = sum(S_none_count + S_low_count + S_high_count + 
                           C_none_count + C_low_count + C_high_count + D_count),
    disease_rate_under5_per_100k = (new_cases_under5 / pop_under5) * 100000,
    .groups = "drop"
  )

ggplot(under5_vax_df, aes(x = time, y = disease_rate_under5_per_100k)) +
  geom_line(color = "#1170AA", linewidth = 1.2) +
  labs(
    title = "Impact of Routine Pediatric Vaccination on Severe Disease (<5 Years)",
    subtitle = "Incidence among children under 5 following burn-in equilibrium",
    x = "Years Since Vaccination Implementation",
    y = "Severe Disease Incidence Rate per 100,000 (<5 yrs)"
  ) +
  theme_minimal(base_size = 12) +
  theme(plot.title = element_text(face = "bold"))
```

```{r tests}
#1. 
library(ggplot2)

ggplot(out_counterfactual$summary, aes(x = year, y = births)) +
  geom_line(color = "#e41a1c", linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Counterfactual: Annual Birth Influx Over Time",
    x = "Year",
    y = "Annual Births"
  )

#2. 
library(dplyr)
library(tidyr)

out_counterfactual$summary |>
  mutate(
    incidence_per_100k = (disease_incidence / N) * 100000
  ) |>
  pivot_longer(
    cols = c(disease_incidence, incidence_per_100k),
    names_to = "metric",
    values_to = "val"
  ) |>
  ggplot(aes(x = year, y = val, color = metric)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ metric, scales = "free_y", labeller = as_labeller(c(
    disease_incidence = "Raw Annual Cases",
    incidence_per_100k = "Incidence Rate (per 100,000)"
  ))) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title = "Counterfactual: Raw Cases vs. Crude Incidence Rate",
    x = "Year", 
    y = "Value"
  )

#3.
# Construct static 1990 demographic matrices by repeating column 1
n_cols <- ncol(mortality_mat)

mortality_mat_static <- matrix(
  mortality_mat[, 1], 
  nrow = nrow(mortality_mat), 
  ncol = n_cols
)

fertility_mat_static <- matrix(
  fertility_mat_scaled[, 1], 
  nrow = nrow(fertility_mat_scaled), 
  ncol = n_cols
)

# Run counterfactual with constant 1990 demography
out_static_counterfactual <- simulate_mcvf_sir_imex_dynamic(
  S0 = S0_eq, C0 = C0_eq, D0 = D0_eq, age = age, time = time_vax, da = da, dt = dt, grid = grid,
  beta_matrix = beta_matrix, age_group_idx = age_group_idx, sigma = sigma,
  susceptibility = susceptibility, infectiousness = infectiousness,
  fertility_matrix = fertility_mat_static,
  mortality_matrix = mortality_mat_static,
  n_immunity = n_immunity, immunity_classes = immunity_classes,
  alpha_L = alpha_L, alpha_H = alpha_H, beta_L = beta_L, beta_H = beta_H,
  wane_high_to_low = wane_high_to_low, wane_low_to_none = wane_low_to_none,
  rho_C = rho_C, rho_I = rho_I,
  frac_maternal_none = frac_maternal_none, frac_maternal_low = frac_maternal_low,
  frac_maternal_high = frac_maternal_high,
  v_rate_none_to_low_matrix = NULL,
  v_rate_to_high_matrix = NULL
)

# Plot comparison between dynamic and static counterfactuals
bind_rows(
  out_counterfactual$summary |> mutate(demographics = "Dynamic (1990-2020)"),
  out_static_counterfactual$summary |> mutate(demographics = "Static (1990 Constant)")
) |>
  ggplot(aes(x = year, y = disease_incidence, color = demographics)) +
  geom_line(linewidth = 1) +
  theme_minimal() +
  labs(
    title = "Counterfactual Cases: Dynamic vs. Static Demographics",
    x = "Year",
    y = "Annual Disease Cases",
    color = "Demographic Setup"
  )
```

```{r dx-two}
library(ggplot2)
library(patchwork)

# 1. Check if total population N is shrinking in static run
p1 <- ggplot(out_static_counterfactual$summary, aes(x = year, y = N)) +
  geom_line(color = "black", linewidth = 1) +
  theme_minimal() +
  labs(title = "Total Population (N)", y = "Count")

# 2. Check if burn-in actually reached steady state at year 20
p2 <- ggplot(out_burnin$summary, aes(x = time, y = disease_incidence)) +
  geom_line(color = "darkblue", linewidth = 1) +
  theme_minimal() +
  labs(title = "Burn-In Trajectory (0-20 yrs)", x = "Burn-in Year", y = "Raw Cases")

# 3. Check Per-Capita Incidence Rate vs Raw Cases
p3 <- out_static_counterfactual$summary |>
  mutate(rate_per_100k = (disease_incidence / N) * 100000) |>
  ggplot(aes(x = year, y = rate_per_100k)) +
  geom_line(color = "darkgreen", linewidth = 1) +
  theme_minimal() +
  labs(title = "Per-Capita Incidence Rate", y = "Cases / 100k")

(p1 | p2) / p3
```

```{r dx-three}
library(dplyr)
library(ggplot2)

# ==============================================================================
# 1. BURN-IN CONVERGENCE CHECK
# Checks if dY/dt reaches equilibrium (~0) at the end of out_burnin
# ==============================================================================
check_burnin_convergence <- function(burnin_obj, state_cols = c("S", "I", "R")) {
  # Extract data frame from list wrapper if out_burnin isn't a plain matrix/data frame
  if (is.list(burnin_obj) && !is.data.frame(burnin_obj)) {
    if ("sim" %in% names(burnin_obj)) {
      df <- as.data.frame(burnin_obj$sim)
    } else if ("out" %in% names(burnin_obj)) {
      df <- as.data.frame(burnin_obj$out)
    } else {
      matrix_idx <- which(sapply(burnin_obj, function(x) is.matrix(x) || is.data.frame(x)))[1]
      if (length(matrix_idx) == 0 || is.na(matrix_idx)) {
        stop("Could not automatically locate simulation matrix inside out_burnin list")
      }
      df <- as.data.frame(burnin_obj[[matrix_idx]])
    }
  } else {
    df <- as.data.frame(burnin_obj)
  }
  
  n_rows <- nrow(df)
  if (is.null(n_rows) || n_rows < 2) stop("Input must have at least 2 time steps")
  
  time_col <- if ("time" %in% names(df)) "time" else if ("year" %in% names(df)) "year" else names(df)[1]
  t_step <- df[[time_col]][n_rows] - df[[time_col]][n_rows - 1]
  
  # Calculate derivatives, summing across stratified sub-compartments if present
  derivs <- sapply(state_cols, function(prefix) {
    matching_cols <- grep(paste0("^", prefix, "($|\\[|_|\\.)"), names(df), value = TRUE)
    if (length(matching_cols) == 0) {
      stop(paste("No columns matching pattern for state:", prefix))
    }
    
    val_curr <- sum(unlist(df[n_rows, matching_cols, drop = FALSE]))
    val_prev <- sum(unlist(df[n_rows - 1, matching_cols, drop = FALSE]))
    
    (val_curr - val_prev) / t_step
  })
  
  cat("=== Burn-In Terminal Derivatives (dState/dt) ===\n")
  print(round(derivs, 5))
  
  unstable <- names(derivs[abs(derivs) > 1e-3])
  if (length(unstable) > 0) {
    message("Warning: Compartments not fully converged: ", paste(unstable, collapse = ", "))
  } else {
    message("System reached equilibrium at end of burn-in")
  }
}

# ==============================================================================
# 2. DEMOGRAPHIC & PROPORTION SHIFT CHECK
# Checks if declining N(t) is altering S/N, I/N, or R/N ratios
# ==============================================================================
check_proportions <- function(sim_obj) {
  df <- as.data.frame(sim_obj) %>%
    mutate(
      N_t = S + I + R,
      prop_S = S / N_t,
      prop_I = I / N_t,
      prop_R = R / N_t
    )
  
  cat("\n=== 2. State Proportions (Start vs End) ===\n")
  cat("Start: Prop S =", round(head(df$prop_S, 1), 5), 
      "| Prop I =", round(head(df$prop_I, 1), 5), "\n")
  cat("End:   Prop S =", round(tail(df$prop_S, 1), 5), 
      "| Prop I =", round(tail(df$prop_I, 1), 5), "\n")
  
  p <- ggplot(df, aes(x = if ("year" %in% names(df)) year else time)) +
    geom_line(aes(y = prop_S, color = "S / N"), linewidth = 1) +
    geom_line(aes(y = prop_I * 1000, color = "I / N (x1000)"), linewidth = 1) +
    labs(
      title = "Compartment Proportions Over Time",
      x = "Time", 
      y = "Proportion",
      color = "Compartment"
    ) +
    theme_minimal()
  
  print(p)
  return(df)
}

# ==============================================================================
# 3. INCIDENCE FORMULATION COMPARISON
# Compares dynamic N(t) denominator vs static initial N denominator
# ==============================================================================
library(dplyr)
library(ggplot2)

check_incidence_scaling <- function(sim_obj, cases_col = "new_cases") {
  df <- as.data.frame(sim_obj)
  
  if (!cases_col %in% names(df)) {
    stop(paste("Cases column", cases_col, "not found in data frame"))
  }
  
  initial_N <- df$S[1] + df$I[1] + df$R[1]
  
  df <- df %>%
    mutate(
      N_t = S + I + R,
      inc_dynamic = (.data[[cases_col]] / N_t) * 100000,
      inc_static  = (.data[[cases_col]] / initial_N) * 100000
    )
  
  cat("\n=== 3. Incidence Rate Comparison ===\n")
  cat("Start - Dynamic N(t):", round(head(df$inc_dynamic, 1), 2), "| Static N0:", round(head(df$inc_static, 1), 2), "\n")
  cat("End   - Dynamic N(t):", round(tail(df$inc_dynamic, 1), 2), "| Static N0:", round(tail(df$inc_static, 1), 2), "\n")
  
  time_var <- if ("year" %in% names(df)) "year" else "time"
  
  p <- ggplot(df, aes(x = .data[[time_var]])) +
    geom_line(aes(y = inc_dynamic, color = "Dynamic N(t) Denominator"), linewidth = 1) +
    geom_line(aes(y = inc_static, color = "Static N0 Denominator"), linetype = "dashed", linewidth = 1) +
    labs(
      title = "Incidence Rate Sensitivity to Population Denominator",
      x = "Time", 
      y = "Cases / 100k",
      color = "Formulation"
    ) +
    theme_minimal()
  
  print(p)
  return(df)
}

# ==============================================================================
# EXECUTION
# ==============================================================================
# 1. Run burn-in check using out_burnin directly
check_burnin_convergence(out_burnin, state_cols = c("S", "I", "R"))

# 2. Run main simulation proportion check (pass your 1990-2020 object name here)
sim_props <- check_proportions(out_vax)

# 3. Run incidence scaling check (update 'new_cases' if your column is named differently)
sim_inc <- check_incidence_scaling(out_main, cases_col = "new_cases")


library(dplyr)
library(ggplot2)

check_burnin_convergence <- function(burnin_list, state_cols = c("S", "C", "D")) {
  df <- as.data.frame(burnin_list$summary)
  n_rows <- nrow(df)
  
  # Keep only columns that exist in the summary table
  valid_cols <- intersect(state_cols, names(df))
  if (length(valid_cols) == 0) {
    stop("None of the specified state_cols exist in out_burnin$summary")
  }
  
  time_col <- if ("time" %in% names(df)) "time" else if ("year" %in% names(df)) "year" else names(df)[1]
  t_step <- df[[time_col]][n_rows] - df[[time_col]][n_rows - 1]
  
  derivs <- sapply(valid_cols, function(col) {
    (df[[col]][n_rows] - df[[col]][n_rows - 1]) / t_step
  })
  
  cat("=== 1. Burn-In Terminal Derivatives (dState/dt) ===\n")
  print(round(derivs, 5))
  
  unstable <- names(derivs[abs(derivs) > 1e-2])
  if (length(unstable) > 0) {
    message("Warning: Compartments still shifting: ", paste(unstable, collapse = ", "))
  } else {
    message("System reached steady-state equilibrium at end of burn-in")
  }
}

# Run burn-in check
check_burnin_convergence(out_burnin, state_cols = c("S", "C", "D"))
```



#pop totals over time
```{r pop-totals}
summary_long <- summary_df |>
  select(time, S, C, D) |>
  pivot_longer(-time, names_to = "compartment", values_to = "persons")

ggplot(summary_long, aes(time, persons, colour = compartment)) +
  geom_line(linewidth = 1.0) +
  scale_colour_manual(values = c(S = "#1b9e77", C = "#d95f02", D = "#7570b3")) +
  labs(
    title = paste0("Total Population Mass Across Compartments (N = ", format(N_target, big.mark=","), ")"),
    x = "Time (years)", y = "Total Persons"
  )

```

#incidence counts
```{r incidence-counts}
summary_aux <- summary_df |>
  select(time, carriage_incidence, disease_incidence, births) |>
  pivot_longer(-time, names_to = "quantity", values_to = "value")

ggplot(summary_aux, aes(time, value)) +
  geom_line(linewidth = 1.0, colour = "#2c7fb8") +
  facet_wrap(~ quantity, scales = "free_y", ncol = 1) +
  labs(
    title = "Annualized Absolute Incidence Counts & Births",
    x = "Time (years)", y = "Persons / Year"
  )

```

#age-specific counts over time
```{r age-counts}
snapshot_counts_long <- snapshots_df |>
  pivot_longer(
    cols = c(S_none_count, S_low_count, S_high_count, 
             C_none_count, C_low_count, C_high_count, D_count),
    names_to = "compartment", values_to = "count"
  ) |>
  mutate(
    compartment = gsub("_count", "", compartment),
    time_label  = paste0("t = ", format(round(time, 1), nsmall = 1))
  )

ggplot(snapshot_counts_long, aes(age, count, colour = compartment)) +
  geom_line(linewidth = 0.9) +
  facet_grid(compartment ~ time_label, scales = "free_y") +
  scale_colour_manual(values = c(
    S_none = "#1b9e77", S_low = "#66c2a5", S_high = "#a6d854",
    C_none = "#fdae61", C_low = "#f46d43", C_high = "#d73027",
    D      = "#7570b3"
  )) +
  labs(
    title = paste0("Age Profiles by Compartment (Counts per ", da, "-Year Bin)"),
    x = "Age (years)", y = "Number of Individuals"
  ) +
  theme(legend.position = "none")

```

#childhood (<5 y-o) invasive disease cases
```{r child-cases}
snapshots_df |>
  filter(age <= 5) |>
  mutate(time_label = paste0("t = ", format(round(time, 1), nsmall = 1))) |>
  ggplot(aes(age, new_disease_cases_per_year, colour = time_label)) +
  geom_line(linewidth = 1.1) +
  labs(
    title = "Annualized New Invasive Hib Cases by Age (<5 Years)",
    x = "Age (years)", y = "New Cases / Year (per Bin)", colour = "Time Step"
  )

```

library(dplyr)
library(ggplot2)

```{r per-100k-incidence}
# 1. Transform summary_df to add per 100k rates
summary_rates_df <- summary_df %>%
   mutate(total_pop = S_none + S_low + S_high + C_none + C_low + C_high + D) %>%
  mutate(
    birth_rate_per_100k    = (births / total_pop) * 100000,
    carriage_rate_per_100k = (carriage_incidence / total_pop) * 100000,
    disease_rate_per_100k  = (disease_incidence / total_pop) * 100000
  )

# 2. Reshape to long format for clean multi-panel ggplotting
library(tidyr)

plot_df <- summary_rates_df %>%
  select(time, birth_rate_per_100k, carriage_rate_per_100k, disease_rate_per_100k) %>%
  pivot_longer(
    cols = -time,
    names_to = "metric",
    values_to = "rate_per_100k"
  ) %>%
  mutate(
    metric = factor(
      metric,
      levels = c("birth_rate_per_100k", "carriage_rate_per_100k", "disease_rate_per_100k"),
      labels = c("Birth Rate (per 100k)", "Carriage Incidence (per 100k)", "Disease Incidence (per 100k)")
    )
  )

ggplot(plot_df, aes(x = time, y = rate_per_100k)) +
  geom_line(color = "#1f77b4", linewidth = 1) +
  facet_wrap(~ metric, ncol = 1, scales = "free_y") +
  scale_x_continuous(breaks = seq(0, 12, by = 2.5)) +
  labs(
    title = "Annualized Incidence Rates per 100,000 Population",
    x = "Time (years)",
    y = "Rate per 100,000 Person-Years"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(size = 11),
    panel.grid.minor = element_blank(),
    plot.title = element_text(hjust = 0.5)
  )
```

# Aggregate Dynamics
These are the total susceptible, infectious, and recovered masses after integrating over age.

```{r aggregate-dynamics}
summary_long <- summary_df |>
  select(time, S, C, D) |>
  pivot_longer(-time, names_to = "compartment", values_to = "mass")

ggplot(summary_long, aes(time, mass, colour = compartment)) +
  geom_line(linewidth = 1.0) +
  scale_colour_manual(values = c(S = "#1b9e77", C = "#d95f02", D = "#7570b3")) +
  labs(
    title = "Aggregate SCD dynamics",
    x = "Time (years)",
    y = "Total mass"
  )
```

```{r incidence-and-births}
summary_aux <- summary_df |>
  select(time, carriage_incidence, disease_incidence, births, mean_lambda) |>
  pivot_longer(-time, names_to = "quantity", values_to = "value")

ggplot(summary_aux, aes(time, value)) +
  geom_line(linewidth = 1.0, colour = "#2c7fb8") +
  facet_wrap(~ quantity, scales = "free_y", ncol = 1) +
  labs(
    title = "Carriage/disease incidence, births, and mean force of infection",
    x = "Time (years)",
    y = NULL
  )
```

# Age Profiles at Selected Times
```{r age-profiles}
snapshot_long <- snapshots_df |>
  pivot_longer(
    cols = c(S_none_count, S_low_count, S_high_count,
             C_none_count, C_low_count, C_high_count, D_count),
    names_to = "compartment", values_to = "density"
  ) |>
  mutate(
    compartment = gsub("_count", "", compartment),
    time_label  = paste0("t = ", format(round(time, 1), nsmall = 1))
  )

ggplot(snapshot_long, aes(age, density, colour = compartment)) +
  geom_line(linewidth = 0.9) +
  facet_grid(compartment ~ time_label, scales = "free_y") +
  scale_colour_manual(values = c(
  S_none = "#1b9e77", S_low = "#66c2a5", S_high = "#a6d854",
    C_none = "#fdae61", C_low = "#f46d43", C_high = "#d73027",
    D = "#7570b3"
)) +
  labs(
    title = "Age profiles of S, C, and D at selected times",
    x = "Age (years)",
    y = "Density"
  ) +
  theme(legend.position = "none")
```
